Designing Types for R, Empirically

### 2020-05-14 paper/main.tex
intro

\begin{document}
\title{Extracting Function Signatures from Program Traces}

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix

@@

>>>>

\begin{document}
\title{Designing Types for R, Empirically}

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix

@@

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks. Unfortunately, intuition may fail when looking at
domain-specific languages as these languages designed for a specific group
of users to solve very specific needs. This is the case of the data science
language R.

R and its ancestor S were designed, implemented, and maintained by
statisticians. Originally they aimed to be glue languages for reading data
and calling statistical routines written in Fortran. Over three decades they
became widely used across many fields for data analysis and visualization.
Modern R, as an object of study, is fascinating. It is a vectorized,
dynamically typed, lazy functional language with limited side-effects,
extensive reflective facilities and retrofitted object-oriented programming
support.

Many of the design decisions that gave us R were intended to foster an
interactive, exploratory, programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcut, and the
automatic conversion between data types.  While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can go
undetected.

Retrofitting a type system to the R programming language would increase our
assurance in the result of data analysis. But, we are faced with two
challenges. First, it is unclear what would be the \emph{right} type system
for a language as baroque as R. For example, one of the most popular data
type, the \code{data.frame}, is manipulated through reflective operations --
a data frame is a table whose columns can be added or removed on the fly.
Second, but just as crucially, designing a type system that will be adopted
would require overcoming some prejudices and educating large numbers of
users.

xxx WIP xxxx

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to propose changes to the
language, and for any changes to be accepted by the user community they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

We propose to capture the degree of polymorphism present in R by means
of a dynamic analysis of widely used libraries. For each function call we
can record the types of its arguments and of its return value. This allows
us to observe how many different combination of types are accepted by any
given function. Unlike many other languages, R has a carefully curated
software repository called CRAN. To be deposited in CRAN, a package must
come with sample dataset, tests and executable use-cases. As part of normal
operations these tests are run regularly and failing packages are removed.
This allowed us to have access to \PACKAGES libraries and about an order of
magnitude more runnable scripts that exercise those libraries.

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item A set of type annotations that captures how programmers use types
in existing R code.
\end{itemize}

One threat to validity of our work is that we rely on dynamic analysis, so
our conclusions are only as good as the coverage of the possibly function
behaviors. Previous work~\cite{issta18} reported that running all the
scripts that come with CRAN packages gives, on average, 68\% test coverage.
We attempted to mitigate the threat coming from the fact that only part of
the code is being exercised by manual analysis. It would be reasonable to
ask for confirmation of the data by static analysis of the code, but sound
static analysis of R is difficult because of the extensive use of reflective
features such as \code{eval} and of the ability to redefine the meaning of
operators such as \code{+} and \code{if}.  Another threat to validity is
that we only have access to code that has been deposited in the CRAN
repository. While this may bias our findings towards code written to be
reusable and, possibly, better engineered than typical user code. This is
also the code that would most benefit from type annotations.
